Concurrent Programming in Harmony

### 2021-07-24 book/paper.tex
Working on an explanation of every type of statement

\chapter{Value Types}
\label{ap:values}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^},
\texttt{and}, \texttt{or}\\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bounded variables, $e$ is a set or dictionary \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{select $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
                            \texttt{S} a list of statements\\
\hline
\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
\hline
\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is the thread-local state \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\end{tabular}

@@

The statement $x = 3$ changes the state by assigning~3 to
variable $x$ (assuming~$x$ was not already~3).
$x$ may be a local variable or a shared variable.
The statement $x = y = 3$ first updates~$y$, then~$x$.
The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = 3$ first computes
the address of $x[\mathtt{f}()]$, then computes the address of
$y[\mathtt{g}()]$, but first assigns to $y[\mathtt{g}()]$ and then
to $x[\mathtt{f}()]$.
The statement $a,b = c$ assumes that $c$ is a tuple with two values.
It first evaluates the addresses of $a$ and $b$ and first assigns to
the latter and then the former.
If $c$ is not a tuple with two values, then Harmony will report an error.

@@

The statement \textbf{atomic $S_1, S_2, ...$} evaluates statements
$S_1, S_2, ...$ atomically in that no other threads can run while
this atomic block is executing.  Typically an atomic block runs
to completion before any other thread can run.  The only exception
to this is if the atomic block executes a \textbf{stop} expression.
In this case, another thread can run.  When the original thread
is resumed (using a \textbf{go} statement), it is once again
atomically executing.

@@

>>>>

\chapter{Value Types and Operators}
\label{ap:values}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
$e$ & $e$ is an expression\\
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\&}, \texttt{|}, \texttt{\^},
\texttt{and}, \texttt{or}\\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{for $a[:b]$ in $e$ [where $c$]: S} & $a$ and $b$ are bounded variables, $e$ is a set or dictionary \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} is a list of statements\\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{invariant e} & \texttt{e} is an invariant\\
\hline
\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{select $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
                            \texttt{S} a list of statements\\
\hline
\texttt{sequential v, ...} & \texttt{v} has sequential consistency \\
\hline
\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is the thread-local state \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\end{tabular}

\section{Single expression evaluation}

Any expression by itself can be used as a statement.  The most common
form of this is a function application, for example: \texttt{f}().
This statement evaluates \texttt{f}() but ignores its result.
It is equivalent to the assignment statement $\_ = \mathtt{f}()$.

@@

The statement $x = 3$ changes the state by assigning~3 to
variable $x$ (assuming~$x$ was not already~3).
$x$ may be a local variable or a shared variable.
The statement $x = y = 3$ first updates~$y$, then~$x$.
The statement $x[\mathtt{f}()] = y[\mathtt{g}()] = \mathtt{h}()$ first computes
the address of $x[\mathtt{f}()]$, then computes the address of
$y[\mathtt{g}()]$, then evaluates $\mathtt{h}()$, then assigns the resulting value to $y[\mathtt{g}()]$ (using its previously computed address), and finally assigns the same value to $x[\mathtt{f}()]$ (again using its previously computed address).
The statement $a,b = c$ assumes that $c$ is a tuple with two values.
It first evaluates the addresses of $a$ and $b$ and first assigns to
the latter and then the former.
If $c$ is not a tuple with two values, then Harmony will report an error.

Assigning to $\_$ evaluates the righthand side expression but is otherwise
a no-op.

@@

The statement \textbf{atomic $S_1; S_2; ...$} evaluates statements
$S_1, S_2, ...$ atomically in that no other threads can run while
this atomic block is executing.  Typically an atomic block runs
to completion before any other thread can run.  The only exception
to this is if the atomic block executes a \textbf{stop} expression.
In this case, another thread can run.  When the original thread
is resumed (using a \textbf{go} statement), it is once again
atomically executing.

@@

\section{\textbf{def}}

The statement \textbf{def} \texttt{m} $a: S_1; S_2: ...$
defines a new program counter constant \texttt{m} referring to
a method that takes an argument $a$ and executes the statements
$S_1, S_2, ...$.  The argument $a$ can be a tuple pattern similar
to those used in \texttt{let} and \texttt{for} statements.
Examples include $()$, $(x,)$, $(x, y)$, and $(x, (y, z))$.
The given local variable names variable names are assigned
upon application.
It is allowed, but discouraged, to updates those local variables
in statements $S_1, S_2, ...$.
Each method has a predefined local variable
\textit{result}, initialized to \texttt{None},
that is returned by the method.  Harmony does
not support a \texttt{return} statement that \emph{breaks out}
of the code before executing the last statement.

\section{\textbf{del}}

The statement \textbf{del} $x$ removes variable $x$ from the
state.  $x$ can be either a local or a shared variable.  For
example, the statement \textbf{del} $x.\texttt{age}$ removes
the \texttt{.age} field from dictionary $x$.
Harmony automatically removes top-level local variables that
are no longer in use
from the state in order to attempt to reduce the number of states
that are evaluated during model checking.

Because Harmony lists are dictionaries, deleting from lists is different
from Python:
$x = [\mathtt{.a}, \mathtt{.b}]; \mathbf{del}~x[0]$ results in $x$ having value
$\{1: \mathtt{.b}\}$ rather than $[\mathtt{.b}]$ (which is $\{0: \mathtt{b}\}$).

\section{\textbf{for ... in [...where]}}

The statement \textbf{for} $x$ \textbf{in} $y: S_1; S_2; ...$
iterates over $y$ and executes for each element the statements
$S_1, S_2, ...$.
which must be a set or a dictionary.  $y$ is evaluated only once at
the beginning of the evaluation of this statement.  In case of a set,
the result is sorted (using Harmony's global order on all values).
For each element, the statements $S_1, S_2, ...$ are executed with
local variable $y$ having the value of the element.
$x$ can be a pattern such as $(a)$ or $(a, (b, c))$.
If the pattern cannot be matched, Harmony detects and error.
It is allowed, but discouraged, to assign different values to $x$
within statements $S_1, S2, ...$.

The statement also supports nesting and filtering.
Nesting is of the form
\textbf{for} $x_1$ \textbf{in} $y_1$ \textbf{for} $x_2$ \textbf{in} $y_2$:
$S_1; S_2; ...$,
which is equivalent to the statement
\textbf{for} $x_1$ \textbf{in} $y_1$: \textbf{for} $x_2$ \textbf{in} $y_2$:
$S_1; S_2; ...$.
Filtering is of the form
\textbf{for} $x$ \textbf{in} $y$ \textbf{where} $z: S_1; S_2; ...$.
For example,
\textbf{for} $x$ \textbf{in} \{ 1 .. 10 \} \textbf{where} $(x \% 2) == 0:
S_1; S_2; ...$ only evaluates statements $S_1, S_2, ...$ for even $x$,
that is, 2, 4, 6, 8, and 10.

\section{\textbf{from ... import}}

\section{\textbf{go}}

\section{\textbf{if ... elif ... else}}

\section{\textbf{import}}

\section{\textbf{invariant}}

\section{\textbf{let}}

\section{\textbf{pass}}

\section{\textbf{select ... in}}

\section{\textbf{sequential}}

\section{\textbf{spawn}}

\section{\textbf{trap}}

\section{\textbf{while}}
